---
title: "PROJECT"
subtitle: "BLAL BLA BLA"
author: anonymous
format:
  html:
    toc: true
    code-tools: true
    code-line-numbers: true
    mainfont: Georgia, serif
    page-layout: article
  pdf:  
    number-sections: true
    code-annotations: none
editor: source
---

:::: {.content-hidden when-format="pdf"}
::: {.callout-warning collapse=false}
 
## Setup

```{r}

```

:::
::::


# Introduction 

- Background
- Problem formulation/scope
- Main modeling idea


# Data Desctription

- Data set
- Histograms
- Correlations
- Has it been used in previous studies?


# Mathematical Model

- Bayesian logistic regression
- Likelihood
- Priors
- Posteriors


# Model Definitions and Implementation

- Tie mathematical model to our implementation
- Ndraws, warmup, etc.
```{r}
# model code
```


# Model Evaluation
- Rhat, ESS, HMC divergences, pp_check(), loo_compare(), classification accuracy
- Prior sensitivity analysis


# Discussion
- Problems, potential improvements?
- Conclusion of analysis

# Lessons Learned


# References

